---
title: "microViz annotated heatmaps"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article will show you how to plot annotated correlation and
microbial composition heatmaps with microViz.

## Setup

```{r setup}
library(corncob)
library(dplyr)
library(phyloseq)
# library(microViz)
```

First we'll get some OTU abundance data from inflammatory bowel disease
patients and controls from the corncob package.

```{r}
data("ibd_phylo", package = "corncob")
ibd_phylo
```

Remove the mostly unclassified species-level data, drop the rare taxa
and fix the taxonomy of the rest. Also drop patients with unclassified
IBD.

```{r}
psq <- ibd_phylo %>% 
  tax_mutate(Species = NULL) %>% 
  tax_filter(min_prevalence = 5) %>% 
  tax_fix() %>% 
  ps_filter(DiseaseState != "IBDundef")
psq
```

## Microbial composition heatmaps

Coming soon.

## Correlation heatmaps

Correlation heatmaps can be a nice way to quickly assess patterns of
associations between numerical variables in your dataset, such as
microbial abundances and other metadata.

Let's make some fake numeric variables to exemplify.

```{r}
psq <- psq %>% 
  ps_arrange(ibd) %>% 
  ps_mutate(
    var1 = rnorm(nsamples(psq), mean = 10, sd = 3),
    var2 = c(
      rnorm(nsamples(psq)*0.75, mean = 4, sd = 2),
      rnorm(1 + nsamples(psq)/4, mean = 9, sd = 3)
      ),
    var3 = runif(nsamples(psq), 2, 10),
    var4 = rnorm(nsamples(psq), mean = 100 + nsamples(psq):0, sd = 20) / 20,
    var5 = rnorm(nsamples(psq), mean = 5, sd = 2),
    var6 = rnbinom(nsamples(psq), size = 1:75/10, mu = 5)
  )
```

### Calculating correlations

The `cor_heatmap` function will correlate all taxa to all numerical sample
data by default.

*Older versions of microViz `cor_heatmap` had a `tax_transform`
argument. But for flexibility, you must now transform your taxa
**before** passing the ps_extra object to `cor_heatmap`.*

```{r}
psq %>% 
  tax_agg("Family") %>% 
  cor_heatmap(vars = c("var1", "var2", "var3", "var4", "var5", "var6"))
```



### Taxon annotations

As seen in the previous plots taxa are annotated by default with 
prevalence and relative abundance.

You can transform the taxa for the abundance annotation. 
The `trans` and `zero_replace` arguments are sent to `tax_transform()`.

```{r, fig.width=7, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  cor_heatmap(
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(ylim = 0:1), 
      Log10. = anno_tax_box(trans = "log10", zero_replace = "halfmin")
    )
  )
```

You can do multiple transformations and or scaling by supplying a function,
that takes a ps_extra or phyloseq object, transforms it, and returns it.

```{r, fig.width=7, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  cor_heatmap(
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Log2 = anno_tax_density(
        joyplot_scale = 1.5, gp = grid::gpar(fill = "black", alpha = 0.2),
        trans = "log2", zero_replace = 1
      ),
      prop_Log2 = anno_tax_density(
        joyplot_scale = 1.5, gp = grid::gpar(fill = "black", alpha = 0.2),
        trans = function(ps) {
          ps %>% 
            tax_transform("compositional", zero_replace = 1) %>% 
            tax_transform("log2", chain = TRUE) 
        }
      )
    )
  )
```






Note that by default the relative abundance is shown only for samples
where the taxon is detected! You can include values for all samples for
all taxa by setting `only_detected = FALSE`.

Let's try this with a heatmap-style density plot annotation. We'll
replace zeroes with ones for an interpretable minimum value on the plot.

We'll compare it side-by-side with the default setting of showing only
distribution of values above the detection threshold.

For zero-inflated microbiome data, showing prevalence and "abundance
when detected" seems like a more informative annotation.

```{r, fig.width=7, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  cor_heatmap(
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(size = grid::unit(10, "mm"), ylim = 0:1), 
      All = anno_tax_density(
        size = grid::unit(20, "mm"),
        trans = "log10", zero_replace = 1, 
        heatmap_colors = viridisLite::turbo(n = 15),
        type = "heatmap", only_detected = FALSE
      ),
      Default = anno_tax_density(
        size = grid::unit(20, "mm"),
        trans = "log10", zero_replace = 1, 
        heatmap_colors = viridisLite::turbo(n = 15),
        type = "heatmap", only_detected = TRUE
      )
    )
  )
```

#### Sorting

By default, rows and columns are sorted using hierarchical clustering
with optimal leaf ordering `"OLO_ward"`. You can use any valid method
from the `seriation` package. You can suppress ordering by using
`seriation_method = "Identity"`. By default this also suppresses column
ordering, so you can set `seriation_method_col = OLO_ward` to keep
ordering.

```{r, fig.width=7, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  tax_sort(by = prev, at = "Family") %>% 
  cor_heatmap(
    seriation_method = "Identity",
    seriation_method_col = "OLO_ward",
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(ylim = 0:1), 
      CLR = anno_tax_box(trans = "clr", zero_replace = "halfmin")
    )
  )
```

#### Taxa annotation side

You can easily put the taxa annotations on another of the heatmap with
e.g. `taxa_side = "left"`

```{r, fig.width=7, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  tax_sort(by = prev, at = "Family") %>% 
  cor_heatmap(
    seriation_method = "Identity",
    seriation_method_col = "OLO_ward",
    taxa_side = "left",
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(ylim = 0:1), 
      CLR = anno_tax_box(trans = "clr", zero_replace = "halfmin")
    )
  )
```

Or on the top or bottom is also possible, this will rotate the heatmap.
Remember to swap the seriation method arguments around!

```{r, fig.width=8, fig.height = 5, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  tax_sort(by = prev, at = "Family") %>% 
  cor_heatmap(
    seriation_method_col = "Identity", #swapped!
    seriation_method = "OLO_ward", #swapped!
    taxa_side = "top",
    vars = paste0("var", 1:6),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(ylim = 0:1), 
      CLR = anno_tax_box(trans = "clr", zero_replace = "halfmin")
    )
  )
```

### Variable annotation

As well as annotating the taxa, you can also annotate the variables.

```{r, fig.width=7, fig.height=8, dpi=120}
psq %>% 
  tax_agg("Family") %>% 
  cor_heatmap(
    vars = paste0("var", 1:6),
    var_anno = varAnnotation(
      Value = anno_var_box(),
      Zscore = anno_var_density(fun = scale, type = "violin")
    )
  )
```

### 
