---
title: "Statistical modelling of individual taxa"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article will give an example of statistical modelling of the
abundances of individual taxa, and visual presentation of the results
using microViz *taxonomic association tree* plots.

## Setup

```{r setup}
library(microViz)
library(corncob)
library(dplyr)
library(ggplot2)
```

First we'll get some OTU abundance data from inflammatory bowel disease
patients and controls from the corncob package.

```{r}
data("ibd_phylo")
ibd_phylo
```

We'll keep only the Ulcerative Colitis and Healthy Control samples, to
simplify the analyses for this example. We'll also remove the Species
rank information, as most OTUs in this dataset are not assigned to a
species. We'll also use `tax_fix` to fill any gaps where the Genus is
unknown, with the family name or whatever higher rank classification is
known.

```{r}
phylo <- ibd_phylo %>% 
  ps_filter(DiseaseState %in% c("UC", "nonIBD")) %>% 
  tax_mutate(Species = NULL) %>% 
  tax_fix()
phylo 
```

Let's have a quick look at the sample data using the `skimr` package.

```{r}
phylo %>% 
  samdat_tbl() %>% 
  dplyr::mutate(across(where(is.character), as.factor)) %>% 
  skimr::skim()
```

Let's make some sample data variables that are easier to use and compare
in the statistical modelling ahead. We will convert dichotomous
categorical variables into similar binary variables (values: 1 for true,
or 0 for false). We will also scale and center the numeric variable for
age.

```{r}
phylo <- phylo %>% 
  ps_mutate(
    UC = ifelse(DiseaseState == "UC", yes = 1, no = 0),
    female = ifelse(gender == "female", yes = 1, no = 0),
    antibiotics = ifelse(abx == "abx", yes = 1, no = 0),
    steroids = ifelse(steroids == "steroids", yes = 1, no = 0),
    age_scaled = scale(age, center = TRUE, scale = TRUE)
  )
```

## Modelling one taxon at a time

### TSS log2 linear regression

We will start by creating a linear regression model for one genus,
*Bacteroides*. We will transform the count data by first making it
proportions, and then taking the logarithm, with base 2. This is what
MaAsLin2 does by default (except they call the compositional
transformation "Total Sum Scaling (TSS)"). We will replace the zeros
with half the minimum observed abundance proportion (of any taxon)
before log2 transformation.

```{r}
parabacteroides_lm <- phylo %>% 
  tax_fix() %>% 
  tax_transform("compositional", rank = "Genus") %>% 
  tax_transform("log2", zero_replace = "halfmin", chain = TRUE) %>% 
  tax_model(
    type = "lm", rank = "Genus", taxa = "Parabacteroides", 
    variables = c("UC", "female", "antibiotics", "steroids", "age_scaled")
  ) 
parabacteroides_lm$Parabacteroides
```

```{r}
summary(parabacteroides_lm$Parabacteroides)
```

This model suggests *Parabacteroides* abundances are significantly lower
in Ulcerative Colitis patients than controls, on average.

### Plotting TSS log2 data

Let's boxplot the transformed data to see what this *Parabacteroides*
association looks like as a crude group difference.

```{r, fig.width=4, dpi=120}
plot_data <- phylo %>% 
  tax_fix() %>% 
  tax_transform("compositional", rank = "Genus") %>% 
  tax_transform("log2", zero_replace = "halfmin", chain = TRUE) %>% 
  ps_get() %>% 
  ps_otu2samdat("Parabacteroides") %>% # adds Parabacteroides as sample data!
  samdat_tbl()

ggplot(plot_data, aes(x = DiseaseState, y = Parabacteroides)) + 
  geom_boxplot(width = 0.5, colour = "grey35") +
  geom_jitter(width = 0.2, alpha = 0.5) + 
  scale_y_continuous(
    breaks = log2(1/2^(0:13)), 
    labels = function(x) paste0(100 * round(2^x, digits = 5), "%"), 
    limits = c(log2(0.00005), log2(0.25))
  ) + 
  theme_bw()
```

### Beta binomial regression

You can also use other regression modelling functions that take a
formula. For example the beta binomial modelling provided in the corncob
package. This approach models both abundance and dispersion, and
directly uses untransformed counts. By default, microViz's `tax_model()`
will use the same formula for both abundance and dispersion modelling,
but you can override this by setting the `phi.formula` argument
yourself. See `vignette("corncob-intro", package = "corncob")` for more
info on these models.

```{r}
parabacteroides_bb <- phylo %>% 
  tax_fix() %>% 
  tax_model(
    type = corncob::bbdml, rank = "Genus", taxa = "Parabacteroides", 
    variables = c("UC", "female", "antibiotics", "steroids", "age_scaled")
  ) 
parabacteroides_bb$Parabacteroides
```

## Model all the taxa!

Now we will fit a similar model for almost every taxon at every rank.
The code for `taxatree_models` is quite similar to `tax_model`. However,
you might need to run `tax_prepend_ranks` to ensure that each taxon at
each rank is always unique. As an example of the problem, Actinobacteria
is the name of both a Phylum and a Class!

```{r}
lm_models <- phylo %>% 
  tax_fix() %>% 
  tax_prepend_ranks() %>% 
  # it makes sense to perform the compositional transformation BEFORE filtering
  tax_transform("compositional", rank = "Genus", keep_counts = TRUE) %>% 
  tax_filter(min_prevalence = 0.1, undetected = 0, use_counts = TRUE) %>% 
  tax_transform(
    trans = "log2", chain = TRUE, zero_replace = "halfmin"
  ) %>% 
  taxatree_models(
    type = lm, 
    ranks = NULL, # uses every rank available except the first
    variables = c("UC", "female", "antibiotics", "steroids", "age_scaled")
  )
```

Why filter the taxa? *It's less likely that we are interested in rare
taxa, and models of rare taxon abundances are more likely to be
unreliable. Reducing the the number of taxa modelled also makes
visualising the results easier!*

```{r}
lm_models
```

### Getting stats from the models

Next we will get a data.frame containing the regression coefficient
estimates, test statistics and corresponding p values from all these
regression models. The function `taxatree_models2stats()` can do this
for any type of model that has a `broom::tidy()` method, as well as for
beta binomial regression models calculated with the `corncob` package
`bbdml()` function.

```{r}
lm_stats <- taxatree_models2stats(lm_models)
lm_stats
```

```{r}
lm_stats$taxatree_stats
```

### Adjusting p values

Using the `taxatree_stats_p_adjust()` function, you can correct for
multiple testing / control the false discovery rate or family-wise error
rate.

Instead of applying these adjustment methods across all 88 taxa models
at all ranks, the default behaviour is to control the family-wise error
rate per rank.

```{r}
lm_stats <- taxatree_stats_p_adjust(
  data = lm_stats, method = "BH", grouping = "rank"
)
# notice the new variable
lm_stats$taxatree_stats
```

### Plot all the taxatree_stats!

`taxatree_plots()` allows you to plot statistics (e.g. point estimates
and significance) from all of the taxa models onto a tree layout. The
taxon models are organised by rank, radiating out from the central root
node from e.g. Phyla around the center to Genera in the outermost ring.

`taxatree_plots()` itself returns a list of plots, which you can arrange
into one figure with the
[`patchwork`](https://patchwork.data-imaginist.com/) package for example
(and/or
[`cowplot`](https://wilkelab.org/cowplot/articles/plot_grid.html)).

```{r, fig.width=7, fig.height=10}
lm_stats %>% taxatree_plots(
  node_size_range = c(1, 3), var_renamer = toupper
) %>% 
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  )
```

#### Taxatree Key

But how do we know which taxa are which nodes? We can create a labelled
grey tree with `taxatree_plotkey`. This labels taxa based on certain
conditions.

```{r, fig.height=5, fig.width=6}
set.seed(123) # label position 
key <- taxatree_plotkey(
  data = lm_stats, 
  taxon_renamer = function(x) stringr::str_remove(x, "[PFG]: "),
  # 2 lines of conditions below, for filtering taxa to be labelled
  rank == "Phylum" | rank == "Genus" & prevalence > 0.25, 
  !grepl("Kingdom", taxon)
) + 
  # add a bit more space for the longer labels by expanding the x axis
  scale_x_continuous(expand = expansion(mult = 0.2))
# all phyla are labelled, and all genera with a prevalence of over 0.2
# except for any taxa whose names (partly) match "Kingdom" 
# (i.e. an unclassified taxon)
key
```

##### Key + Trees

Let's put the key and some of the trees together in one `patchwork`
figure. Getting the sizing right on these combined plots can be very
tricky! Pay attention to the absolute height and width of the plot
output.

`gridExtra::grid.arrange()` or `cowplot::plot_grid()` are alternatives
you can also try. `cowplot::get_legend()` can be particularly useful.

```{r, fig.width=13, fig.height=5.5, dpi=120}
trees <- lm_stats %>% 
  taxatree_plots(node_size_range = c(1, 2.25)) %>%
  .[1:4] %>% 
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  )

panel <- patchwork::wrap_plots(key, trees, nrow = 1, widths = 8:7)
set.seed(111)
panel
```

You could save the plot with `ggsave()` like this.

```{r eval=FALSE}
set.seed(111)
ggsave("test.png", panel, width = 13, height = 5.5, dpi = 120, device = "png")
```

##### Alternative label styling

You can change the default styling of the labels by first suppressing
the automatic drawing of labels with `.draw_label = FALSE` in
`taxatree_plotkey()` and then adding your own custom-style labels with
`taxatree_plot_labels()`. Here we will draw some yellow labels.

```{r, fig.height=5, fig.width=7}
taxatree_plotkey(
  data = lm_stats, .draw_label = FALSE,
  rank %in% c("Phylum", "Family") & !grepl("Bacteria", taxon),
  prevalence > 0.2 | rank == "Phylum"
) %>% 
  taxatree_plot_labels(
    taxon_renamer = function(x) stringr::str_remove(x, "[PFGO]: "),
    # default fun is ggrepel::geom_text_repel
    fun = ggrepel::geom_label_repel, 
    # arguments modifying label style
    size = 2.5, alpha = 0.5, colour = "black", fill = "gold1", 
    label.size = 0.05, label.r = unit(0.05, "lines"), 
    label.padding = unit(0.15, "lines"), segment.size = 0.5,
    # arguments affecting label positioning
    box.padding = 0.05, x_nudge = 0.4, y_nudge = 0.05,
    hjust = 0.5, seed = 123
  )
```

#### Directly labelling taxa

You can directly label taxatree_plots too, but it is better to only do
this for a few taxa. You must run `taxatree_label()` first to create a
"label" indicator variable.

```{r, fig.width=6.5, fig.height=4}
lm_stats %>% 
  taxatree_label(
    rank == "Genus", p.value < 0.05 | prevalence > 0.5, estimate > 0
  ) %>% 
  taxatree_plots() %>% 
  .[[1]] %>% # show only the first plot
  taxatree_plot_labels(
    taxon_renamer = function(x) stringr::str_remove(x, "G: "),
    fun = ggrepel::geom_label_repel, x_nudge = 0.7, hjust = 0.5, size = 2
  ) +
  # the lines below allow expansion and make some space for labels
  coord_fixed(clip = "off", expand = TRUE) +
  scale_x_continuous(expand = expansion(mult = c(0.15, 0.25)))
```

#### Sorting taxa nodes

If you like, you can sort the nodes by sorting the taxa in the ps_extra
object.

```{r, fig.width=5, fig.height=4}
lm_stats %>% 
  tax_sort(by = "prev", at = "Genus") %>%
  taxatree_plots() %>% 
  .[[1]] # show only the first plot
```

You can chain multiple `tax_sort()` calls together to fine-tune the
order of the nodes on the tree to your own preference.

```{r, fig.width=5, fig.height=4}
lm_stats %>% 
  tax_sort(by = "prev", at = "Family") %>%
  tax_sort(by = "name", at = "Phylum") %>%
  tax_sort(by = "rev") %>%
  taxatree_plots() %>% 
  .[[1]] # show only the first plot
```

#### Plotting adjusted p values

Remember we made adjusted p values earlier? Let's plot those instead.
Just to show how it's done, we'll also change the symbol used to
identify the significant sites to a cross, and we'll also relax the
significance threshold to 0.1.

It looks like only the disease state (having ulcerative colitis) shows
any significant associations after this FDR correction.

```{r, fig.width=7, fig.height=6}
lm_stats %>% 
  taxatree_plots(
    sig_stat = "p.adj.BH.rank", sig_threshold = 0.1, 
    sig_shape = "cross", sig_size = 1.5,
    node_size_range = c(1, 3), var_renamer = toupper
) %>% 
  .[1:4] %>% # keep only first 4 plots
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  )
```

## Beta-binomial regression example

The corncob package provides beta-binomial regression models. See the
paper [here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7514055/), and
the helpful package vignette:
`vignette("corncob-intro", package = "corncob")`.

We will filter the taxa more strictly (by a higher prevalence threshold)
before this type of modelling. We do not need to transform the data, as
this approach uses counts.

```{r}
bb_models <- phylo %>% 
  tax_fix() %>% 
  tax_prepend_ranks() %>% 
  tax_filter(min_prevalence = 0.3) %>% 
  taxatree_models(
    type = corncob::bbdml, 
    ranks = c("Phylum", "Class", "Order", "Family"),
    variables = c("UC", "female", "antibiotics", "steroids", "age_scaled")
  )
bb_models
```

When extracting stats from corncob beta-binomial models, you need to
specify which parameter estimate you want, "mu" for differential
abundance, or "phi" for differential variability or overdispersion.

```{r}
bb_stats <- taxatree_models2stats(bb_models, param = "mu")
bb_stats
bb_stats$taxatree_stats
```

```{r, fig.width=5, fig.height=5}
bb_stats %>% taxatree_plots(
  node_size_range = c(1, 4), colour_trans = "identity"
) %>% 
  # keep only first 4 plots
  .[1:4] %>% 
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  ) 
```

### Alternative layouts

You do not need to make circular tree plots if you don't want to!

```{r, fig.width=5, fig.height=5}
alt_trees <- bb_stats %>% 
  taxatree_plots(
    node_size_range = c(1, 4), circular = FALSE, colour_trans = "identity"
  ) %>% 
  # keep only first 4 plots
  .[1:4] %>% 
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  ) & # & is used by patchwork to modify multiple ggplots (instead of +)
  coord_flip() & 
  scale_y_reverse()

alt_trees
```

Let's add the key for this layout and label it manually with
`taxatree_plot_labels()`.

```{r, fig.width=5, fig.height=8}
alt_tree_key <- taxatree_plotkey(
  data = bb_stats, circular = FALSE, .draw_label = FALSE,
  rank == "Family"
) %>% 
  taxatree_plot_labels(
    circular = FALSE, 
    taxon_renamer = function(x) stringr::str_remove(x, "[PFG]: "),
    hjust = 0.5, force = 0, nudge_y = 2, size = 3
  )  + 
  coord_flip() + 
  scale_y_reverse(expand = expansion(mult = c(0.1, 1))) +
  theme(plot.title = element_text(size = 14))

patchwork::wrap_plots(alt_tree_key, alt_trees, nrow = 2, heights = 1:2)
```

You don't have to use a regular tree!

Alternative layouts from the igraph package are possible, such as the
Kamada and Kawai spring algorithm ("kk") or Fruchterman and Reingold
force-directed algorithm ("fr"). You must set a layout_seed number for
these layouts to ensure they are always the same.

```{r, fig.width=5, fig.height=5}
bb_stats %>% taxatree_plots(
  node_size_range = c(1, 4), 
  colour_trans = "identity", layout = "kk", layout_seed = 321
) %>% 
  # keep only first 4 plots
  .[1:4] %>% 
  patchwork::wrap_plots(
    ncol = 2, guides = "collect"
  )
```

# Session info

```{r}
devtools::session_info()
```
